---
title: "Starting with data"
format: html
editor_options: 
  chunk_output_type: console
---

## Loading the survey data

```{r}
download.file(
  url = "https://ndownloader.figshare.com/files/2292169",
  destfile = "portal_data_joined.csv"
)
```

## Load package

```{r}
library(tidyverse)
```

## Load and view data

```{r}
surveys <- read_csv("portal_data_joined.csv")
```

```{r}
print(surveys) # default 10 rows
surveys
print(surveys, n = 50)
```


```{r}
head(surveys) # default 6 rows
head(surveys, n = 15)
```

```{r}
tail(surveys) # default 6 rows
tail(surveys, n = 15)
```

```{r}
str(surveys)
```

```{r}
glimpse(surveys)
```

```{r}
view(surveys) # Even better: use F2 shortcut
```


## Challenge: View data

Based on the output of str(surveys), can you answer the following questions?

- What is the class of the object surveys?
- How many rows and how many columns are in this object?

```{r}
class(surveys)
```


## Indexing and subsetting data frames ----

```{r}
surveys[1, 1] # First row and first column, output: dataframe

surveys[1, 6] # First row and sixth column, output: dataframe

surveys[1, ] # Entire first row, output: dataframe

surveys[, 1] # Entire first column, output: dataframe

surveys[1] # Entire first column, output: dataframe

surveys[c(1, 2, 3), c(5, 6)] # Frist 3 rows and fifth/sixth columns, output: dataframe
surveys[1:3, 5:6]
```


```{r}
surveys[1:6, ] # First six rows
head(surveys)

nrow(surveys)
surveys[34781:34786,]
surveys[34781:nrow(surveys),]
tail(surveys)

surveys[[1]] # First row, output: vector

surveys[[1, 1]] # First column and first row, output: vector
```


```{r}
surveys[, -1]
surveys[-(7:nrow(surveys)),]
```


```{r}
surveys["species_id"]
surveys[, "species_id"]

surveys[["species_id"]]

surveys$species_id
```


## Challenge: Subsetting datafarme

Create a data.frame (surveys_200) containing only the data in row 200 of the surveys dataset.

Notice how nrow() gave you the number of rows in a data.frame?

Use that number to pull out just that last row from the surveys dataset.
Compare that with what you see as the last row using tail() to make sure it’s meeting expectations.
Pull out that last row using nrow() instead of the row number.
Create a new data frame (surveys_last) from that last row.
Use nrow() to extract the row that is in the middle of the data frame. Store the content of this row in an object named surveys_middle.

Combine nrow() with the - notation above to reproduce the behavior of head(surveys), keeping just the first through 6th rows of the surveys dataset.